\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-09 concepts.tex
cleaned up section on TokenStates, moving to section on Rules now.

  \item \textit {duration}: The constraint \textit{start} $+$
    \textit{duration} = \textit{end} is enforced automatically.

@@

  \item \textit{state}: Tokens can be \texttt{ACTIVE, INACTIVE,
      MERGED}, or \texttt{REJECTED}. The state variable captures the
    token's current state and its reachable states through further
    restriction.  This is \eus mechanism to support the CAIP approach.

@@

@@

% \textit{Objects (continued)}

% As we have noted, Tokens describe some aspect of an Object in
% time. Objects thus may have many Tokens in a plan in order to describe
% their state and behavior throughout all points in time of the
% plan. Within this general framework, we note a few particulars:

% Static Facts: Classes without predicates lead to Objects without
% Tokens. This arises where an objects state or behavior is independent
% of time. For example, a domain may have a set of locations and/or
% paths for which there is nothing more to say than that they exist.

% Timelines: Often objects in a domain must be described by exactly one
% token for every given timepoint in the plan. Such objects are so
% common that we provide a special construct in \eu to extend these
% semantics to derived classes. Any instances of a class derived from a
% Timeline will induce ordering requirements among its tokens in order
% to ensure no temporal overlap may occur among them. See below for
% further discussion.

% Resources: Metric resources, e.g. the energy of a battery or the
% capacity of a cargo hold, are objects with an explicit quantitative
% state in time and with a circumscribed range of changes that can occur
% to impact that state i.e. produce, consume, use, change. These changes
% are captured as tokens. Resources are such a common requirement for
% \eu users that special constructs are also provided for
% them. Instances of classes derived from a Resource will induce
% ordering requirements on their Tokens in order to ensure that the
% level of the resource remains within specified limits.

\item[\textbf{Timelines}] It may be sufficient to maintain only a
  partial-order among tokens. However, it is often the case that
  tokens represent states and actions of a single object in the
  system. Such tokens are typically mutually-exclusive. \eu uses a
  Timeline structure \cite{mus94} to concisely capture system
  components whose behavior is described over time in this manner.

@@

For example, consider the tire-world domain. The tire was located in
the trunk with the predicate tireLocated(Trunk) and located on the
ground with the predicate tireLocated(Ground). Clearly, the same tire
cannot be in both places at once. A Timeline provides a simple method
for aggregating the statements about a tire such that they are
mutually exclusive.

Figure 1: A Timeline for a Tire

Figure 1 illustrates how a Timeline can be used to describe the
whereabouts of a tire. The tire is an object in the tire-world
represented as a timeline. It has predicates associated with it which
can describe its states and actions over time. The predicate names
need not contain the tire prefix; this is now implicit since the
tokens are assigned to a specific tire instance (i.e. an instance of a
Timeline). While Timelines are a useful element of the \eu planning
paradigm, they are not essential. A more general notion of a system
Object can be used which does not impose restrictions of
mutual-exclusion or non-zero duration. This can be important in
supporting partial-order planning.

For this example, the state of the tire is specified using 3
contiguous tokens. The end and start time-points are related by an
equality constraint. A Moving predicate has been introduced to cover
the transition from one location to another. It takes 2 location
arguments. Notice that precedence relationships exist between tokens
such that they cannot overlap but the start and end times may remain
flexible. To illustrate this, sample times are included. Assume a
total time range of interest between 0 and 1000. In addition, tokens
on Timelines have a minimum duration of 1. As a result the values
shown are the most flexible possible values for the start and end of
each token. There are a number of advantages of allowing this
flexibility. First, the basic structure of the plan can be developed
without over-committing to specific times. If it is not necessary for
the tire to be on the ground at time-step 4, then a planner should not
be forced to specify it. Such an approach permits a least-commitment
approach to planning. Second, in many domains it is simply impossible
to know in planning exactly how long an activity might take. For
example, a common activity of driving a car from one point to another
on a road can take varying amounts of time depending on traffic and
traffic lights. In such circumstances it is more practical to express
upper and lower bounds on durations which naturally lead to intervals
for start and end times. In such domains, flexibility in planning aids
robustness in execution.  Rules

In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.  Token State Model

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid. There are 2 possibilities to
consider:

The slave is inserted as an active token in the plan. As such, rules
may be activated on the slave, and it may consume resources.

The token is merged with a matching token already in the plan. Once a
slave is merged, the requirement it represents are considered
satisfied. The process of merging passes on all restrictions imposed
on the slave to the active token upon which it is merged. Merging a
token requires finding a target active token that is compatible with
the inactive token. For an active token and an inactive token to be
compatible requires that they are instances of the same predicate and
that no intersections between corresponding variables are empty. The
effect of merging is illustrated in Figure 3.

Figure 3: Merging an inactive token on an active token. Domain
restrictions occur in highlighted variables of active token.

On creation of a token, where the commitment has not been made yet to
activate or merge the token, the token is said to be inactive. There
is a nuance to the state model for tokens which relate to the mode of
its creation. If the token is allocated explicitly by an external
actor, rather than internally through rule firing, it may include the
state rejected indicating the planner is permitted to reject the
token. A valid plan can include rejected tokens. This typically arises
where the token represents a goal that is preferable to achieve but
not mandatory. This state is not reachable for slaves since that would
imply selective adherence to the domain model. Figure 4 presents the
state transition diagram for token states relevant for planning. The
transitions are operations on a partial plan which can decide an
outcome for an inactive token. As operations which may arise in
search, they must be reversible during backtracking. The cancellation
operations in each case are also shown.

Each of the token variables, including the parameter variables, has a
domain of values assigned to it.  The variables may also participate
in constraints that specify which value combinations are valid.  To
allow the specification of multiple values, e.g, to express a range of
possible start times, variables are used to specify parameter, start
and end time values for a token.  As a result, a token $T$ is a tuple
$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
$s\leq e$) are variables encoded as flexible token start and end time
points. Consequently, partial plans represent a \textit{range of
  possible solutions} rather than an explicit definition of a single
trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
of a part of such plans with tokens representing constraints on
flexible temporal intervals. Systematic computation is reflected by
the use of the \texttt{Allen Algebra} \cite{allen84} temporal
relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
interval arithmetic in the \eu domain model that allows
encoding of temporal constraint relationships between tokens on the
same or other concurrent timelines. Details of propagation algorithms
or of interval logic are beyond the scope of this paper.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
\caption{\small Timeline evolution can be described by simple finite
  state transitions where temporal relationships describe the arc
  transition and temporal extant of state are token
  durations. Constraints here are for \emph{meets} or its equivalent
  inverse \emph{met\_by}.}
\label{fig:FSM}
\vskip-0.3cm
\end{figure}

In describing the evolution of state via timelines, transitions within
a timeline can be described by simple finite state
machines. Fig. \ref{fig:FSM} for example describes a timeline
evolution for an abstract timeline entity describing navigation. Arc
transitions represent temporal constraints and token duration are
represented by temporal extant of state. The entire vehicle can then
be modeled as a number of state transitions with concurrency for each
of the modeled state variables defined a priori. However this simple
view is often complicated when temporal constraints between state
variables are articulated in our domain models. If there is a
criticism of this approach, modeling then, requires a proper balance
between abstraction, state variable description and concurrency
described by the inter-connectedness between such finite state
machines.

@@

% \textit{Summary}

% This has presented the main elements of the EUROPA planning approach. They are:

% \begin{enumerate}

% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
% states and actions in time. Note that tokens do not discriminate
% between state and action.

% \item Token States which are the basis of planning operations on a partial
% plan.

% \item Constraints to describe relationships among tokens. This provides an
% expressive method of describing interactions among tokens in the
% context of Temporal Planning.

% \item Timelines as a concise abstraction to express the evolution of state
% and behavior for a system component. It provides semantics of mutual
% exclusion sequencing in time. Other core abstractions are available in
% \eu for handling metric resources.

% \item Domain rules to describe internal and external relationships on and
% between tokens respectively. Rules are applied to active tokens
% (master tokens) referred to as masters and typically produce inactive
% tokens (slaves) which are a key vehicle through which the planning
% process evolves. The rule structure in \eu differs from the more
% restrictive commitment to preconditions and effects in classical
% planning which prohibits durative actions and disembodied effects
% (i.e. effects which may not occur until some temporal distance after
% the end of the action).

>>>>

  \item \textit {duration}: The duration of the temporal extent over which the predicate is defined. The constraint \textit{start} $+$
    \textit{duration} = \textit{end} is enforced automatically.

@@

  \item \textit{state}: Tokens can be \texttt{ACTIVE, INACTIVE,
      MERGED}, or \texttt{REJECTED}. The state variable captures the
    token's current state and its reachable states through further
    restriction.  This is required to support CAIP's approach to planning where goals and subgoals can be satisfied by either adding new intervals (Tokens) to the plan, or by merging with existing intervals. Token State lifecycle is examined in detailed in a separate section below.

@@

\item[\textbf{Built-in Object Types}] In a domain model, there may be some classes that don't have any time-dependent state or behavior, and therefore those classes will not have any Token types associated with them, for instance in the Shopping Agent model, the set locations and products are static for the agent's purposes so we may only want to say what they are but not have any state or behavior associated with them.

However, the most common case in any non-trivial domain model is that Objects will have many associated Tokens in order to describe their state and behavior throughout a plan. There are a couple of Object Types that are so commonly used in domain descriptions, that \eu provides a built-in implementation for them:

\begin{enumerate}
    \item \textit{Timelines}: Often objects in a domain must be described by exactly one
    token for every given timepoint in the plan. Any instances of a class derived from \eu's built-in
    Timeline class will induce ordering requirements among its tokens
     to ensure no temporal overlap may occur among them  \cite{mus94}. 
     \comment{add small example?}

    \item \textit{Resources}: Metric resources, e.g. the energy of a battery or the
    capacity of a cargo hold, are objects with an explicit quantitative
    state in time and with a circumscribed range of changes that can occur
     to impact that state i.e. produce, consume, use, change. 
     Resources are such a common requirement for \eu users that built-in object types (with their corresponding token types to denote production, consumption, etc) are provided for them. Instances of classes derived from a Resource will induce
     ordering requirements on their Tokens in order to ensure that the
     level of the resource remains within specified limits.
     \comment{add small example?}
\end{enumerate}

\item[\textbf{Token State Model}] In \eu's representation, goals are temporally scoped states that need to be achieved, or actions that must be performed. As a result, \eu internally represents goals as Tokens.  Goals can be posted explicitly, for instance, when a user asks the Shopping Agent to own a specific product by a specific time, or implicitly when subgoals are created as a result of domain compatibilities (described in the CAIP framework above), for instance, when the Shopping Agent must get to a Location where a Product is available before it is able to purchase it, in this case the original goal of purchasing the product will spawn a subgoal of being at a specific location.

Looking at a how a simple Shopping Agent goal would be addresed by \eu helps illustrate how Token States support the planning process.

Let's assume we start with a partial plan that places the Shopping agent at home at time 0 and with no Products in its possession, then we post a goal for the Shopping Agent to own a Drill at time 10.

\comment{Add figure showing an AgentLocation timeline with an At(Home) token with start=0 and end open, an empty AgentBag timeline, and a Own(Drill) token with start=10 floating below both timelines}

Notice that the \texttt{Own(Drill)} token that represents the goal is not immediately placed on the \texttt{AgentBag} timeline since \eu must determine whether it can achieve that goal and how. When the goal is posted and before any decisions are made on it, the  \texttt{Own(Drill)} token is created in an INACTIVE state, then \eu looks at the current partial plan and considers 2 alternatives:

\begin{enumerate}
    \item If there are any tokens in the plan that are compatible with the new goal, \eu may attempt to satisfy the goal by merging it with any of them (if there is more than one it can try them one at a time), in this case, the state of the token would change to MERGED.
    \item Instead of merging with existing tokens in the plan, \eu may decide to insert the new token into the plan, in that case its state would change to ACTIVE.
\end{enumerate}

When a token is merged, all restrictions imposed
on the merged token are transferred to the active token upon which it is merged. Merging a
token requires finding a target active token that is compatible with
the inactive token. Two tokens are merge-compatible if they are instances of the same predicate and
that no intersections between corresponding variables are empty. The
effect of merging is illustrated in Figure 3.

\comment{Add figure showing the effect of merging}

When a token is activated all compatibilities associated with its token type are evaluated, which may lead to new subgoals being generated. In our example, there should be a compatibility stating that the Shopping Agent must be at the same location as the product it is purchasing. In that case, activating the \texttt{Own(Drill)} token will cause a new  \texttt{At(Drill.location)} to be created in an INACTIVE state, which in turn will have to be dealt with.

\comment{Add figure showing the effect of activating the token by putting it on the AgentBag timeline, and a new At(Drill.location) token floating below}

If neither activating nor merging works, a token may become REJECTED, this is possible only for goals posted by the end user and specified as optional. Subgoals generated as a result of token activation cannot be rejected, if a subgoal cannot be satisfied through activation or merging then it will cause the original activation decision that spawned it to be retracted (\eu's search mechanisms are explained in detail in a separate section later on).

As we can see, a token's state can go from INACTIVE to ACTIVE, MERGED or REJECTED, this lifecycle is illustrated in the figure below.

\comment{Add figure showing possible state transitions}

@@

\item[\textbf{Rules}] In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid.

@@

\comment{instead of summary, tie every concept to the shopping example?}

This section has presented the main elements of \eu's planning representation:

\begin{enumerate}
    \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
    states and actions in time. Note that tokens do not discriminate between state and action.

    \item Token States which are the basis of planning operations on a partial plan.

    \item Constraints to describe relationships among tokens. This provides an
    expressive method of describing interactions among tokens in the context of Temporal Planning.

    \item Timelines as a concise abstraction to express the evolution of state
    and behavior for a system component. It provides semantics of mutual
    exclusion sequencing in time. Other core abstractions are available in
    \eu for handling metric resources.

    \item Domain rules to describe internal and external relationships on and
     between tokens respectively. Rules are applied to active tokens (master tokens) 
     referred to as masters and typically produce inactive
     tokens (slaves) which are a key vehicle through which the planning
     process evolves. The rule structure in \eu differs from the more
     restrictive commitment to preconditions and effects in classical
     planning which prohibits durative actions and disembodied effects
     (i.e. effects which may not occur until some temporal distance after
     the end of the action).
\end{enumerate}

%Each of the token variables, including the parameter variables, has a
%domain of values assigned to it.  The variables may also participate
%in constraints that specify which value combinations are valid.  To
%allow the specification of multiple values, e.g, to express a range of
%possible start times, variables are used to specify parameter, start
%and end time values for a token.  As a result, a token $T$ is a tuple
%$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
%variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
%$s\leq e$) are variables encoded as flexible token start and end time
%points. Consequently, partial plans represent a \textit{range of
%  possible solutions} rather than an explicit definition of a single
%trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
%of a part of such plans with tokens representing constraints on
%flexible temporal intervals. Systematic computation is reflected by
%the use of the \texttt{Allen Algebra} \cite{allen84} temporal
%relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
%interval arithmetic in the \eu domain model that allows
%encoding of temporal constraint relationships between tokens on the
%same or other concurrent timelines. Details of propagation algorithms
%or of interval logic are beyond the scope of this paper.
%
%\begin{figure}
%\centering
%\includegraphics[scale=0.3]{figs/FSM-transition.pdf}
%\caption{\small Timeline evolution can be described by simple finite
%  state transitions where temporal relationships describe the arc
%  transition and temporal extant of state are token
%  durations. Constraints here are for \emph{meets} or its equivalent
%  inverse \emph{met\_by}.}
%\label{fig:FSM}
%\vskip-0.3cm
%\end{figure}
%
%In describing the evolution of state via timelines, transitions within
%a timeline can be described by simple finite state
%machines. Fig. \ref{fig:FSM} for example describes a timeline
%evolution for an abstract timeline entity describing navigation. Arc
%transitions represent temporal constraints and token duration are
%represented by temporal extant of state. The entire vehicle can then
%be modeled as a number of state transitions with concurrency for each
%of the modeled state variables defined a priori. However this simple
%view is often complicated when temporal constraints between state
%variables are articulated in our domain models. If there is a
%criticism of this approach, modeling then, requires a proper balance
%between abstraction, state variable description and concurrency
%described by the inter-connectedness between such finite state
%machines.

%For example, consider the tire-world domain. The tire was located in
%the trunk with the predicate tireLocated(Trunk) and located on the
%ground with the predicate tireLocated(Ground). Clearly, the same tire
%cannot be in both places at once. A Timeline provides a simple method
%for aggregating the statements about a tire such that they are
%mutually exclusive.
%
%Figure 1: A Timeline for a Tire
%
%Figure 1 illustrates how a Timeline can be used to describe the
%whereabouts of a tire. The tire is an object in the tire-world
%represented as a timeline. It has predicates associated with it which
%can describe its states and actions over time. The predicate names
%need not contain the tire prefix; this is now implicit since the
%tokens are assigned to a specific tire instance (i.e. an instance of a
%Timeline). While Timelines are a useful element of the \eu planning
%paradigm, they are not essential. A more general notion of a system
%Object can be used which does not impose restrictions of
%mutual-exclusion or non-zero duration. This can be important in
%supporting partial-order planning.
%
%For this example, the state of the tire is specified using 3
%contiguous tokens. The end and start time-points are related by an
%equality constraint. A Moving predicate has been introduced to cover
%the transition from one location to another. It takes 2 location
%arguments. Notice that precedence relationships exist between tokens
%such that they cannot overlap but the start and end times may remain
%flexible. To illustrate this, sample times are included. Assume a
%total time range of interest between 0 and 1000. In addition, tokens
%on Timelines have a minimum duration of 1. As a result the values
%shown are the most flexible possible values for the start and end of
%each token. There are a number of advantages of allowing this
%flexibility. First, the basic structure of the plan can be developed
%without over-committing to specific times. If it is not necessary for
%the tire to be on the ground at time-step 4, then a planner should not
%be forced to specify it. Such an approach permits a least-commitment
%approach to planning. Second, in many domains it is simply impossible
%to know in planning exactly how long an activity might take. For
%example, a common activity of driving a car from one point to another
%on a road can take varying amounts of time depending on traffic and
%traffic lights. In such circumstances it is more practical to express
%upper and lower bounds on durations which naturally lead to intervals
%for start and end times. In such domains, flexibility in planning aids
%robustness in execution.

@@
